---
title: "Adam theme"
author: "Adam Shen"
output:
  html_document:
    css: css/base.css
    highlight: themes/adam.theme
    math_method: katex
    md_extensions: -smart
  pdf_document:
    latex_engine: xelatex
    highlight: themes/adam.theme
    md_extensions: -smart
    includes:
      in_header: header.tex
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

# Packages and setup

```{r}
library(tidyverse)

set.seed(20)
```

# The `rowwise` function

The [`rowwise`](https://dplyr.tidyverse.org/reference/rowwise.html) function allows for the grouping
of data by rows in order to perform some operation across the values found in its columns.

Some basic data:

```{r}
nums <- tibble(
  x1 = sample(1:5, size=6, replace=TRUE),
  x2 = sample(1:5, size=6, replace=TRUE),
  x3 = sample(1:5, size=6, replace=TRUE),
  x4 = sample(1:5, size=6, replace=TRUE)
)

nums
```

I won't actually be using `x4` in the demo below, but I'm including it to make the data a bit more
realistic. Oftentimes, you will have more variables in your data set than the ones you are trying
to operate on, i.e. you can't always just `tidyselect::everything()`!

## Using `c_across` with `rowwise`

The [`c_across`](https://dplyr.tidyverse.org/reference/c_across.html) function is often used with
`rowwise` in order to **combine values across columns**. What happens if we use `c_across` without
`rowwise`?

```{r}
nums %>%
  mutate(test = mean(c_across(x1:x3)))
```

It can be seen that the `test` column has a single value that is repeated throughout. Without
grouping the data rowwise, the mean is computed using **all** the values across `x1`, `x2`, and
`x3`:

```{r}
mean(c(nums$x1, nums$x2, nums$x3))
```

Now, if we group the data rowwise before computing the mean across `x1`, `x2`, and `x3`:

```{r}
nums %>%
  rowwise() %>%
  mutate(test = mean(c_across(x1:x3)))
```

We can check that the results are what we wanted:

```{r}
mean(c(3, 1, 4))
mean(c(5, 2, 4))
```

They match &#8212; great!

**Note:** `rowwise` is a type of grouping. From the first two lines of the tibble output above, we
can see that after creating the `test` variable, the data is still grouped (rowwise)! Don't forget
to `ungroup` when you're done mutating.

## Rowwise matrix operations

### Output is a vector of length 1

Suppose you wanted to perform the following computation using the values found in each row:

\[\mathbf{x}'\mathbf{A}\mathbf{x}\]

We will first create a function that:

1. Takes the combined values over a selection of columns (i.e. the result of `c_across`)
2. Converts the values to a matrix (a $n \times 1$ column vector)
3. Performs the matrix operation and reduces the resulting $1 \times 1$ matrix to a vector of
length 1

\

```{r}
quadratic <- function(x, A) {
  x <- as.matrix(x)
  
  drop(t(x) %*% A %*% x)
}
```

For the $\mathbf{A}$ matrix, let's use:

```{r}
A <- matrix(
  sample(0:5, size=9, replace=TRUE),
  nrow=3, ncol=3
)

A
```

Putting it all together:

```{r}
nums %>%
  rowwise() %>%
  mutate(test = quadratic(c_across(x1:x3), A)) %>%
  ungroup()
```

Checking our work:

```{r}
quadratic(c(3, 1, 4), A)
quadratic(c(5, 2, 4), A)
```

### Output is a vector of length greater than 1

Suppose you wanted to perform the following computation using the values found in each row:

\[\mathbf{A}\mathbf{x}\]

and wanted each value to go into its own column. This can be accomplished by first creating a
function similar to the previous, but returning a list containing a named vector instead. Then we
can use [`unnest_wider`](https://tidyr.tidyverse.org/reference/hoist.html) to unnest the values
within the list into their own columns.

Creating the right-multiplying function:

```{r}
right_mult <- function(x, A) {
  x <- as.matrix(x)
  
  drop(A %*% x) %>%
    set_names(., paste0("new_x", 1:length(.))) %>%
    list()
}
```

Creating the new list-column:

```{r}
nums %>%
  rowwise() %>%
  mutate(test = right_mult(c_across(x1:x3), A)) %>%
  ungroup()
```

Unnesting the list contents into their own columns:

```{r}
nums %>%
  rowwise() %>%
  mutate(test = right_mult(c_across(x1:x3), A)) %>%
  ungroup() %>%
  unnest_wider(test)
```

Checking our work:

```{r}
right_mult(c(3, 1, 4), A)
right_mult(c(5, 2, 4), A)
```
